Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-01 Manuscript/02-Method.Rmd
ENAR Submission Version

To encourage proper smoothing of the functions, we adopt the idea of smoothing penalties from smoothing spline models. 
<!-- The basic idea is to set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}), in the prior density function. However, the direct integration of smoothing penalty with sparsity penalty is not obvious. -->
While it is hard to directly integrate smoothing penalty with sparsity penalty, @Marra2011 proposed a reparameterization to implement smoothing penalty in the design matrix. Given the smoothing penalty matrix $\bs S_j$ is symmetric and positive semi-definite for univariate smoothing functions, we apply eigendecomposition on the penalty matrix $\bs S = \bs U \bs D \bs U^T$ , where eigenvectors and eigenvalues are in the matrices $\bs U$ and $\bs D$ respectively. The zero eigenvalue and its corresponding eigenvector span the linear space of the smoothing function, which allows us to separate the linear space from the smoothing function. By multiplying the design matrix $X$ and eigenvector matrix $U$ and proper scaling with eigenvalues, we can have a new design matrix such that the smoothing function of variable $x_j$ can be written as
$$
B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \beta_j X^0_j + \bs {\beta_j^\tp}^T \bs X_j^\tp ,
$$
where $\bs X_j \equiv \begin{bmatrix} X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable; the coefficients $\bs \beta_j \equiv \begin{bmatrix} \beta_j : \bs \beta^\tp_j \end{bmatrix}$ is an augmentation of the coefficient scalar $\beta_j$ of linear space and the coefficient vector $\bs \beta^\tp_j$ of non-linear space.

@@

The family of spike-and-slab (SS) priors regression models dominates Bayesian high-dimensional analysis for its utility in outcome prediction and variable selection. 
<!-- We defer to Bai et al. [@Bai2021Review] for an in-depth introduction to spike-and-slab priors. To summarize, spike-and-slab priors are a family of mixture distributions that comprises a skinny spike density $f_{\text{spike}}(\cdot)$ for weak signals and a flat slab density $f_{\text{slab}}(\cdot)$ for strong signals, mathematically -->
<!-- $$ -->
<!--  \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta). -->
<!-- $$ -->
<!-- The most distinct feature of SS priors is that it is conditioned on a latent binary variable $\gamma \in \{0,1\}$ that indicates whether the variable $x$ is included in the model. There are various spike-and-slab priors depending on the choice for the spike density $f_{\text{spike}}(\cdot)$ and the slab density $f_{\text{slab}}(\cdot)$, see  -->
The development includes @George1993; @George1997; @Chipman1996 for grouped variables; @Brown1998 for multivariate outcomes; @Ishwaran2005; @Clyde2004; @Rockova2014a for optimization-based model fitting algorithms and reference therein.
<!-- Comments -->
<!-- Even the conjugate normal prior of $\bs \beta$ provides analytic simplicity, there is no closed form solution for the posterior density of $\bs \gamma$. @George1993 described a Gibbs sampling algorithm to approximate the posterior distribution of $\bs \gamma$ for variable selection. The same authors later compared SSVS with other Bayesian variable selection approaches, providing some empirical suggestions on the advantages and disadvantages of various formulations and estimation algorithms. [@George1997] Among the various problem SSVS addresses, high-dimensionality is not one of them, due to the intensive computation needs of MCMC algorithms. @George1997 suggested MCMC algorithms works well for medium size of predictors (p=25), which is not feasible for high-dimensional data, where the number of predictors easily exceeds 100. -->
<!-- However, the major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018].  -->
Particularly, the development of spike-and-slab Lasso (SSL) [@Rockova2018b; @Rockova2018] substantially improves the scalability of SS models by using coordinate-wise implementations for posterior exploration. Yi and his group later developed generalized and cox SSL models for high-dimensional -omics data analysis.[@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The SSL prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{align} \label{eq:ssl}
\beta | \gamma &\sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber\\
\text{or} \qquad \beta | \gamma &\sim DE(0, S), 0 < s_0 < s_1,
\end{align}
<!-- Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density, -->
<!-- It can be aslso expressed in the form of a congregated double exponential distribution -->
<!-- \begin{equation} \label{eq:ssl} -->
<!-- \end{equation} -->
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$.

@@

<!-- ### Spike-and-Slab Lasso Prior -->
We introduce a novel SSL type prior of GAMs for high-dimensional nonlinear modeling with bi-level selection. 
<!-- The proposed prior extends from the spike-and-slab lasso prior described in Equation (\ref{eq:ssl}). Given the re-parameterized deign matrix $\bs X_j = \begin{bmatrix} X^0_j : \bs X_j^\tp \end{bmatrix}$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j = \begin{bmatrix} \beta_j : \bs \beta_j^\tp \end{bmatrix}$.  -->
Specifically, we respectively impose independent group priors on the linear space coefficient $\beta_j$ and on the nonlinear space coefficients $\bs \beta^\tp_j$ of $K_j$ elements, 
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j
\end{align}
where $\gamma_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}$ are two latent indicator variables, indicating if the model includes the linear effect and the nonlinear effect of the $j$th variable respectively. $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}.
<!-- To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], as the $\beta_j$ and $\bs \beta^\tp_j$ have different indicator variables $\gamma_j$, $\gamma_j^\tp$ respectively. It is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^\tp_j$.  This converges to the SB-GAM [@Bai2021]. Conversely, it is also possible to relax the assumption such that each coefficient $\beta_{jk} \in \bs \beta^\tp_j$ has its own latent indicator $\gamma_{jk}$, but at the cost of complicating the bi-level function selection. This reduces the proposed prior to the classic spike-and-slab lasso prior. -->
<!-- The reparameterization introduced in Section \ref{sec:BHAM} grants the validity of the proposed prior. First of all, the smoothing function bases are linear dependent and ask for extra attention. The eigeondecomposition remedies the problem and hence our prior can be set to be conditionally independent. Secondly, the eigenvalue scaling provides a panacea to allow unified scale parameters for all bases of all smoothing functions of variables.  -->

The rest of the hierarchical prior follows the traditional spike-and-slab lasso prior: we set up hyper-priors on $\gamma_j, \gamma^\tp_j$ to allow local adaption of the shrinkage using a Bernoulli distribution. The two indicators of the $j$th predictor, $\gamma_j$ and $\gamma^\tp_j$, shares the same probability parameter $\theta_j$,
$$
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^\tp | \theta_j \sim Bin(1, \theta_j).
\end{aligned}
$$
This is to leverage the fact that the probability of selecting the bases of a smoothing function should be similar, while allowing different penalty on the linear space and non-linear space of the smoothing function. The hyper prior of $\gamma_{j}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1), i.e. $a = 1, b = 1$. When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM, and visually presented in Figure \ref{fig:SSprior}.

\begin{figure}
\centering
\caption{Directed acyclic grpah of BHAM-SSL model with parameter expansion. Elliposes are stochastic nodes, rectangles and are deterministic nodes. }
\label{fig:SSprior}
\begin{tikzpicture} [
staticCompo/.style = {rectangle, minimum width=1cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
outCome/.style={ellipse, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
mymatrix/.style={matrix of nodes, nodes=outCome, row sep=1em},
PriorBoarder/.style={rectangle, minimum width=5cm, minimum height=10cm, text centered, fill=lightgray!30},
background/.style={rectangle, fill=gray!10,inner sep=0.2cm, rounded corners=5mm}
]

@@

\end{tikzpicture}
\end{figure}

@@

In this section, we introduce the two EM-based algorithms, EM-Coordinate Descent (EM-CD) and EM-Iterative Weighted Least Squares (EM-IWLS) algorithms, to fit BHAMs. The two proposed algorithms provides different utilities: the EM-CD algorithm is specifically to fit BHAM with an expedited computation, recommended to use in high and ultra-high dimensional setting. A specific concern of EM-CD algorithm is that it provides no information for inference. In contrast, the EM-IWLS can extract the variance-covariance matrix and be used to construct Wald type inference.
<!-- Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS priors but also Student's t-priors and double exponential priors. To note, SB-GAM[@Bai2020; @Bai2021] also used an EM-CD algorithm. The main difference between proposed EM-CD algorithm and that in SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior, while the proposed prior is pairwise independent requiring no special treatment in the CD algorithm. -->
<!-- , stemming from previous EM-coordinate descent algorithms[@Tang2018; @Tang2019] and EM-IWLS [@Rockova2014a]. Instead of simulating the posterior distribution via MCMC, we iteratively estimate the MAP. Compared to the EM-IRLS algorithm, the EM-coordinate descent provides an expedited model fitting process at the cost of model uncertainty.  -->
<!-- The main difference between Bai's work and the proposed models are two-folded: 1) the proposed models are feasible to answer the question if the nonlinear effects are necessary for the predictors, while Bai's work can't due to the "all-in-all-out" approach; 2) the proposed models can provide uncertainly measures in addition to the point estimates of the coefficients, while Bai's work can't not due to the model fitting algorithm. One of the exception is Bai (in progress), who a grouped spike-and-slab LASSO prior [@Rockova2018] to address the grouping structure of spline bases.  -->
<!-- ### EM algorithms -->

@@

We use the EM algorithm to find the MAP estimate of $\Theta$. Since the latent binary indicators $\bs \gamma$ are not of our primary interest, we treat them as the "missing data" in the EM algorithm. Naturally, in the E-step, we calculate the expectation of posterior density function of $\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$ with respect to the latent indicators $\bs \gamma$ conditioning on the values from previous iteration $\Theta^{(t-1)}$, 
$$
E_{\bs \gamma|\Theta^{(t-1)}}\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) .
$$
Hereafter, we use the shorthand notation $E(\cdot)\equiv E_{\bs \gamma|\Theta^{(t-1)}}(\cdot)$. In the M-step, we find the $\Theta^{(t)}$ that maximize $E\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$. The E- and M- steps are iterated until the algorithm converge.

To note here, the log-posterior density of BHAMs (up to additive constants) can be written as a two-part equation
$$ \log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) = Q_1(\bs \beta, \phi) + Q_2 (\bs \gamma,\bs \theta),$$
where
$$ Q_1\equiv Q_1(\bs \beta, \phi) = \log f(\textbf{y}|\bs \beta, \phi) + \sum\limits_{j=1}^p\left[\log f(\beta_j|\gamma_j)+\sum\limits_{k=1}^{K_j} \log f(\beta^{\tp}_{jk}|\gamma^{\tp}_{jk})\right]$$ and 
$$Q_2 \equiv Q_2(\bs\gamma,\bs\theta) = \sum\limits_{j=1}^{p} \left[ (\gamma_j+\gamma_{j}^{\tp})\log \theta_j + (2-\gamma_j-\gamma_{j}^{\tp}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log f(\theta_j).$$
$Q_1$ and $Q_2$ are respectively the log posterior density of the coefficients $\bs \beta$ and the log posterior density of the probability parameters $\bs \theta$ conditioning on $\bs \gamma$. Meanwhile, conditioning on $\bs \gamma$, $Q_1$ and $Q_2$ are independent and can be maximized separately for $\bs \beta, \phi$ and $\bs \theta$. In the optimization step, $Q_1$ can be treated as penalized likelihood function and solved via CD algorithm or IWLS algorithm. Maximization of $E(Q_2)$ can be solved via closed form equations following the beta-binomial conjugate relationship.

>>>>

To encourage proper smoothing of the functions, we adopt the idea of smoothing penalties from smoothing spline models [@Wood2017].
<!-- The basic idea is to set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}), in the prior density function. However, the direct integration of smoothing penalty with sparsity penalty is not obvious. -->
While it is hard to directly integrate smoothing penalty with sparsity penalty, @Marra2011 proposed a reparameterization to implement smoothing penalty in the design matrix. Given the smoothing penalty matrix $\bs S_j$ is symmetric and positive semi-definite for univariate smoothing functions, we apply eigendecomposition on the penalty matrix $\bs S = \bs U \bs D \bs U^T$ , where eigenvectors and eigenvalues are arranged in the matrices $\bs U$ and $\bs D$ respectively. The zero eigenvalue and its corresponding eigenvector span the linear space of the smoothing function, which allows us to separate the linear space from the smoothing function. By multiplying the design matrix $X$ and eigenvector matrix $U$ and properly scaling with eigenvalues, we can have a new design matrix such that the smoothing function of variable $x_j$ can be written as
$$
B_j(x_j) = B_j^0(x_j) + B_j^\tp (x_j) = \beta_j X^0_j + \bs {\beta_j^\tp}^T \bs X_j^\tp ,
$$
where $\bs X_j \equiv \begin{bmatrix} X_j^0 : \bs X_j^{\tp} \end{bmatrix}$ as the basis function matrix for the $j$th variable; the coefficients $\bs \beta_j \equiv \begin{bmatrix} \beta_j : \bs \beta^\tp_j \end{bmatrix}$ is an augmentation of the coefficient scalar $\beta_j$ of linear space and the coefficient vector $\bs \beta^\tp_j$ of non-linear space.

@@

The family of spike-and-slab (SS) priors regression models dominates Bayesian high-dimensional analysis for its utility in outcome prediction and variable selection. 
<!-- We defer to Bai et al. [@Bai2021Review] for an in-depth introduction to spike-and-slab priors. To summarize, spike-and-slab priors are a family of mixture distributions that comprises a skinny spike density $f_{\text{spike}}(\cdot)$ for weak signals and a flat slab density $f_{\text{slab}}(\cdot)$ for strong signals, mathematically -->
<!-- $$ -->
<!--  \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta). -->
<!-- $$ -->
<!-- The most distinct feature of SS priors is that it is conditioned on a latent binary variable $\gamma \in \{0,1\}$ that indicates whether the variable $x$ is included in the model. There are various spike-and-slab priors depending on the choice for the spike density $f_{\text{spike}}(\cdot)$ and the slab density $f_{\text{slab}}(\cdot)$, see  -->
The development includes @George1993; @George1997; @Chipman1996 for grouped variables; @Brown1998 for multivariate outcomes; @Ishwaran2005; @Clyde2004; @Rockova2014a for optimization-based model fitting algorithms and reference therein.
<!-- Comments -->
<!-- Even the conjugate normal prior of $\bs \beta$ provides analytic simplicity, there is no closed form solution for the posterior density of $\bs \gamma$. @George1993 described a Gibbs sampling algorithm to approximate the posterior distribution of $\bs \gamma$ for variable selection. The same authors later compared SSVS with other Bayesian variable selection approaches, providing some empirical suggestions on the advantages and disadvantages of various formulations and estimation algorithms. [@George1997] Among the various problem SSVS addresses, high-dimensionality is not one of them, due to the intensive computation needs of MCMC algorithms. @George1997 suggested MCMC algorithms works well for medium size of predictors (p=25), which is not feasible for high-dimensional data, where the number of predictors easily exceeds 100. -->
<!-- However, the major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018].  -->
Particularly, the development of spike-and-slab Lasso (SSL) [@Rockova2018b; @Rockova2018] substantially improves the scalability of SS models by using coordinate-wise implementations for posterior exploration. Yi and his group later developed generalized and cox SSL models for high-dimensional -omics data analysis.[@Tang2017a; @Tang2017; @Tang2018; @Tang2019] The SSL prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{align} \label{eq:ssl}
\beta | \gamma &\sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber\\
\text{or} \qquad \beta | \gamma &\sim DE(0, S), 0 < s_0 < s_1,
\end{align}
<!-- Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density, -->
<!-- It can be aslso expressed in the form of a congregated double exponential distribution -->
<!-- \begin{equation} \label{eq:ssl} -->
<!-- \end{equation} -->
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$.

@@

<!-- ### Spike-and-Slab Lasso Prior -->
We introduce a novel SSL type prior of GAMs for high-dimensional nonlinear modeling with bi-level selection. 
<!-- The proposed prior extends from the spike-and-slab lasso prior described in Equation (\ref{eq:ssl}). Given the re-parameterized deign matrix $\bs X_j = \begin{bmatrix} X^0_j : \bs X_j^\tp \end{bmatrix}$ for the $j$th variable, we impose a two-part SSL prior to the coefficients $\bs \beta_j = \begin{bmatrix} \beta_j : \bs \beta_j^\tp \end{bmatrix}$.  -->
Specifically, we impose independent group priors on the linear space coefficient $\beta_j$ and on the nonlinear space coefficients $\bs \beta^\tp_j$ of $K_j$ elements, 
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^\tp_{jk} | \gamma^\tp_{j},s_0,s_1 &\simiid DE(0,(1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1), k=1,\dots, K_j
\end{align}
where $\gamma_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}$ are two latent indicator variables, indicating if the model includes the linear and nonlinear effects of the $j$th variable respectively. $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}.
<!-- To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], as the $\beta_j$ and $\bs \beta^\tp_j$ have different indicator variables $\gamma_j$, $\gamma_j^\tp$ respectively. It is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^\tp_j$.  This converges to the SB-GAM [@Bai2021]. Conversely, it is also possible to relax the assumption such that each coefficient $\beta_{jk} \in \bs \beta^\tp_j$ has its own latent indicator $\gamma_{jk}$, but at the cost of complicating the bi-level function selection. This reduces the proposed prior to the classic spike-and-slab lasso prior. -->
<!-- The reparameterization introduced in Section \ref{sec:BHAM} grants the validity of the proposed prior. First of all, the smoothing function bases are linear dependent and ask for extra attention. The eigeondecomposition remedies the problem and hence our prior can be set to be conditionally independent. Secondly, the eigenvalue scaling provides a panacea to allow unified scale parameters for all bases of all smoothing functions of variables.  -->

The rest of the hierarchical prior follows the traditional spike-and-slab lasso prior: we set up hyper-priors on $\gamma_j, \gamma^\tp_j$ to allow local adaption of the shrinkage using a Bernoulli distribution. The two indicators of the $j$th predictor, $\gamma_j$ and $\gamma^\tp_j$, shares the same probability parameter $\theta_j$,
$$
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^\tp | \theta_j \sim Bin(1, \theta_j).
\end{aligned}
$$
This is to leverage the fact that the probability of selecting the bases of a smoothing function should be similar, while allowing different penalty on the linear space and non-linear space of the smoothing function. The hyper prior of $\gamma_{j}$ decides the sparsity of the model at the function selection level, while that of $\gamma_{j}^\tp$ decides the smoothness of the spline function at basis function level. Meanwhile, we specify that $\gamma_{j}$ and $\gamma_{j}^\tp$ are independently distributed for analytic simplicity. The discussion on how to improve the hierarchy of $\gamma_{j}$ and $\gamma_{j}^\tp$ is located in Section \ref{sec:concl}. We further specify the parameter $\theta_j$ follows a beta distribution with given shape parameters $a$ and $b$,
$$
\theta_j \sim Beta(a, b).
$$
The beta distribution is a conjugate prior for the binomial distribution and hence provides some computation convenience. For simplicity, we focus on a special case of beta distribution, uniform (0,1), i.e. $a = 1, b = 1$. When the variable have large effects in any of the bases, the parameter $\theta_j$ will be estimated large, which in turn encourages the model to include the rest of bases, achieving the local adaption among spline bases. Hereafter, we refer Bayesian hierarchical generalized additive models with the spike-and-slab lasso prior as the BHAM, and is visually presented in Figure \ref{fig:SSprior}.

\begin{figure}
\centering
\caption{Directed acyclic grpah of BHAM model with parameter expansion. Elliposes are stochastic nodes, rectangles and are deterministic nodes. }
\label{fig:SSprior}
\scalebox{0.8}{%
\begin{tikzpicture} [
staticCompo/.style = {rectangle, minimum width=1cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
outCome/.style={ellipse, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=blue!30},
mymatrix/.style={matrix of nodes, nodes=outCome, row sep=1em},
PriorBoarder/.style={rectangle, minimum width=5cm, minimum height=10cm, text centered, fill=lightgray!30},
background/.style={rectangle, fill=gray!10,inner sep=0.2cm, rounded corners=5mm}
]

@@

\end{tikzpicture}
}
\end{figure}

@@

In this section, we introduce the two EM-based algorithms, EM-Coordinate Descent (EM-CD) and EM-Iterative Weighted Least Squares (EM-IWLS) algorithms, to fit BHAMs. The two proposed algorithms provides different utilities: the EM-CD algorithm is specifically to fit BHAM with an expedited computation, recommended to use in high and ultra-high dimensional setting. A shortcoming of EM-CD algorithm is that it provides no information for uncertainty inference. In contrast, the EM-IWLS can extract the variance-covariance matrix of the coefficients and be used to construct Wald type inference.
<!-- Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS priors but also Student's t-priors and double exponential priors. To note, SB-GAM[@Bai2020; @Bai2021] also used an EM-CD algorithm. The main difference between proposed EM-CD algorithm and that in SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior, while the proposed prior is pairwise independent requiring no special treatment in the CD algorithm. -->
<!-- , stemming from previous EM-coordinate descent algorithms[@Tang2018; @Tang2019] and EM-IWLS [@Rockova2014a]. Instead of simulating the posterior distribution via MCMC, we iteratively estimate the MAP. Compared to the EM-IRLS algorithm, the EM-coordinate descent provides an expedited model fitting process at the cost of model uncertainty.  -->
<!-- The main difference between Bai's work and the proposed models are two-folded: 1) the proposed models are feasible to answer the question if the nonlinear effects are necessary for the predictors, while Bai's work can't due to the "all-in-all-out" approach; 2) the proposed models can provide uncertainly measures in addition to the point estimates of the coefficients, while Bai's work can't not due to the model fitting algorithm. One of the exception is Bai (in progress), who a grouped spike-and-slab LASSO prior [@Rockova2018] to address the grouping structure of spline bases.  -->
<!-- ### EM algorithms -->

@@

As the latent binary indicators $\bs \gamma$ are not of our primary interest, we treat them as the "missing data" in the EM algorithm. In the E-step, we calculate the expectation of posterior density function of $\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$ with respect to the latent indicators $\bs \gamma$ conditioning on the values from previous iteration $\Theta^{(t-1)}$, 
$$
E_{\bs \gamma|\Theta^{(t-1)}}\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) .
$$
Hereafter, we use the shorthand notation $E(\cdot)\equiv E_{\bs \gamma|\Theta^{(t-1)}}(\cdot)$. In the M-step, we find the $\Theta^{(t)}$ that maximize $E\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X})$. The E- and M- steps are iterated until the algorithm converges.

To note here, the log-posterior density of BHAMs (up to additive constants) can be written as a two-part equation
$$ \log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) = Q_1(\bs \beta, \phi) + Q_2 (\bs \gamma,\bs \theta),$$
where
$$ Q_1\equiv Q_1(\bs \beta, \phi) = \log f(\textbf{y}|\bs \beta, \phi) + \sum\limits_{j=1}^p\left[\log f(\beta_j|\gamma_j)+\sum\limits_{k=1}^{K_j} \log f(\beta^{\tp}_{jk}|\gamma^{\tp}_{jk})\right]$$ and 
$$Q_2 \equiv Q_2(\bs\gamma,\bs\theta) = \sum\limits_{j=1}^{p} \left[ (\gamma_j+\gamma_{j}^{\tp})\log \theta_j + (2-\gamma_j-\gamma_{j}^{\tp}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log f(\theta_j).$$
$Q_1$ and $Q_2$ are respectively the log posterior density of the coefficients $\bs \beta$ and the log posterior density of the probability parameters $\bs \theta$ conditioning on $\bs \gamma$. They are conditionally independent and can be maximized separately for $\bs \beta, \phi$ and $\bs \theta$. In the optimization step, $E(Q_1)$ can be treated as penalized likelihood function and solved via CD algorithm or IWLS algorithm. Maximization of $E(Q_2)$ can be solved via closed form equations following the beta-binomial conjugate relationship.

<!-- ### EM-Coordinate Descent -->
<!-- \label{sec:EMCD} -->
<!-- \vspace{\baselineskip} -->
\textit{EM-Coordinate Descent\label{sec:EMCD}} When the prior distribution of the coefficients is set to mixture double exponential, coordinate descent algorithm can be used to estimate the parameters in the M-step. Coordinate descent [@Friedman2010] is an optimization algorithm that offers extreme computational advantage, and famous for its application in optimizing the $l_1$ penalized likelihood function. To recall, the density function of spike-and-slab mixture double exponential prior can be written as
$$
f(\beta|\gamma, s_0, s_1) = \frac{1}{2\left[(1-\gamma)s_0 + \gamma s_1\right]}\exp(-\frac{|\beta|}{(1-\gamma)s_0 + \gamma s_1}),
$$
and $E(Q_1)$ can be expressed as a $l_1$ penalized likelihood function 
\begin{equation}\label{eq:Q1_CD}
E(Q_1) = \log f(\textbf{y}|\bs \beta, \phi) - \sum\limits_{j=1}^p\left[E({S_j}^{-1})|\beta_j|+\sum\limits_{k=1}^{K_j}E({S^{\tp}}^{-1}_{j})|\beta_{jk}|\right],
\end{equation}
where $S_{j} = (1-\gamma^{0}_{j}) s_0 + \gamma^{0}_{j} s_1$ and $S^\tp_{j} = (1-\gamma^\tp_{j}) s_0 + \gamma^\tp_{j} s_1$. To calculate two unknown quantities $E({S_j}^{-1})$ and $E({S^\tp}^{-1}_j)$, the posterior probability $p_{j} \equiv \pr(\gamma^{0}_{j}=1|\Theta^{(t-1)})$ and $p_{j}^\tp \equiv \pr(\gamma^\tp_{j}=1|\Theta^{(t-1)})$ are necessary, which can be derived via Bayes' theorem. The calculation of $p_j^\tp$ is slightly different from that of $p_j$, as $p_j^\tp$ depends on the value of the vector $\bs \beta^\tp_{j}$ and $p_j$ only depends on the scalar $\beta_j$. The calculation follows the equations below,
\begin{align*}
p_{j} &= \frac{\pr(\gamma_{j} = 1|\theta_j)f(\beta_{j}|\gamma_{j}=1, s_1) }{\pr(\gamma_{j} = 1|\theta_j)f(\beta_{j}|\gamma_{j}=1, s_1) + \pr(\gamma_{j} = 0|\theta_j)f(\beta_{j}|\gamma_{j}=0, s_0)}\\
p^\tp_{j} &= \frac{\pr(\gamma^{\tp}_{j} = 1|\theta_j)\prod\limits_{k=1}^{K_j}f(\beta_{jk}|\gamma^{\tp}_{j}=1, s_1) }{\pr(\gamma^{\tp}_{j} = 1|\theta_j)\prod\limits_{k=1}^{K_j}f(\beta_{jk}|\gamma^{\tp}_{j}=1, s_1) + \pr(\gamma^{\tp}_{j} = 0|\theta_j)\prod\limits_{k=1}^{K_j}f(\beta_{jk}|\gamma^{\tp}_{j}=0, s_0)}
\end{align*}
where  $\pr(\gamma_{j}^{0} = 1|\theta_j) = \pr(\gamma_{j}^\tp = 1|\theta_j) = \theta_j$, $\pr(\gamma_{j}^{0} = 0|\theta_j) = \pr(\gamma_{j}^\tp = 0|\theta_j) = 1-\theta_j$, $f(\beta|\gamma=1, s_1) = \text{DE}(\beta|0 , s_1)$, $f(\beta|\gamma=0, s_0) = \text{DE}(\beta|0 , s_0)$. It is trivial to show
\begin{align}\label{eq:exp_scale}
&E(\gamma_{j})  = p_{j} & &E(\gamma^{\tp}_{j}) = p_{j}^{\tp}\nonumber\\
&E({S}^{-1}_{j}) = \frac{1-p_{j}}{s_0} + \frac{p_{j}}{s_1} & &E({S^\tp}^{-1}_{j}) = \frac{1-p_{j}^{\tp}}{s_0} + \frac{p_{j}^{\tp}}{s_1}.
\end{align}
After substituting the calculated quantities, $E(Q_1)$ can be seen as a $l_1$ penalized likelihood function with the regularization parameter $\lambda = E(S^{-1})$, and hence be optimized via coordinate descent algorithm. Independently, the remaining parameters of interest $\bs \theta$ can be updated by maximizing $E(Q_2)$. As the beta distribution is a conjugate prior for Bernoulli distribution, $\bs \theta$ can be easily updated with a closed form equation,
\begin{equation}\label{eq:update_theta}
\theta_j = \frac{p_j + p^\tp_{j} + a - 1 }{a + b}.
\end{equation}
